---
title: 'Makeover monday: Price of a night out'
author: Felipe Marin
date: '2018-11-26'
slug: makeover-monday-price-of-a-night-out
categories:
  - Visualisation
tags: [R, plotly, MakeoverMonday]
description: ''
featured_image: ''
---
```{r, echo=FALSE}
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=6, warning = FALSE, message=FALSE)
```

It was my first participation in [#MakeoverMonday](https://www.makeovermonday.co.uk/), 
a weekly workshop where participants submit different visualisations from a dataset 
available for that week. It's a great initiative
and a way to discover and practice new ways of communicating insights from data. 

Usually the submissions are made in [Tableau Public](https://public.tableau.com/) as the 
authors of the blog are "Tableau Zen Masters".
I have used 
Tableau a few times, and it's certainly in my wish list to learn, 
but since I have a bit of practice and it's more relevant for
work 
I will be using [plotly in R](https://plot.ly/r/) to generate these visuals, 
although it make time some more time as there
is more coding involved. 

This week the theme was ['The cost of a night out'](https://www.thrillist.com/news/nation/cost-of-a-night-out-cities-around-the-world) 
from selected cities
around the world. 
The original plot show the cost of a 'club' night out, itemising costs of club entry, a 'Big Mac',
taxi and two drinks.  The costs vary greatly, from a highest average cost of $93 in Zurich, while the lowest 
was in Mexico city where a party night for two costs $20 on average.

The data that could be used included also the average costs of a 'date night' out. I wanted to include both 
costs in the visual, which I will generate in plotly. To start, let's load the libraries

```{r}
#load libraries
library(plotly)
library(readxl)
library(dplyr)
library(magrittr)
```

```{r, include = FALSE}
# Sys.setenv("plotly_username"="fmarinp")
# Sys.setenv("plotly_api_key"="Uxyx970oW9MktIffPNVn")

```

I  downloaded  the data, but using the package `data.world` it is possible to grab the data directly 
from the web.

```{r, echo=FALSE}
#get data
file_location <- "/Users/fmarinp/Downloads/Cost of a night out around the world.xlsx"
ubs_data <- read_excel(file_location, sheet = 1 )
```

The plot I have in mind is called ["tornado diagram"](https://en.wikipedia.org/wiki/Tornado_diagram), where
I will the itemised costs of a night out for the different cities.

The 'tornado' look is given by sorting the total costs in descending order. I
do some basic manipulation in `dplyr` to get the total cost by category and 
city, so I can put the city with the most expensive date night on top:

```{r}
#get ranking of cities by cost of date night
rank_cities <- ubs_data %>%
  filter(Category == 'Date night') %>%
  group_by(City) %>% 
  summarise(totalCost = sum(Cost)) %>%
  arrange(totalCost) %>%
  select(City) %>% pull()

#City as factor sorted by date night cost
ubs_data$City  = factor(ubs_data$City, levels = rank_cities)

#order items as well.
ubs_data$Item  = factor(ubs_data$Item, 
                        levels = c("Dinner" ,"Drinks at Dinner",
                                   "Cinema entry", "Taxi", 
                                   "Big Mac","Club entry", 
                                   "2 Longdrinks"))
```


I will use `plotly` to generate the figure. First, I organise my data frame in a way that I can plot
both date nights and party nights at the same level. In this case, the party nights will be on the 
negative side of x

```{r}
dataPlot <- ubs_data %>% 
  mutate(MyCost = ifelse(Category == 'Date night', Cost, -Cost)) 

#this makes the ticks on the negative side to show positive sign
tick_vals = c(-100,-50,0,50,100,150,200)
tick_labels = c('100','50','0','50','100','150','200')
```


For the color palette, I choose shades of blue for the date night, shades of red for party night
using [`colorRampPalette`](https://moderndata.plot.ly/create-colorful-graphs-in-r-with-rcolorbrewer-and-plotly/)
```{r}
#date night
b <- colorRampPalette(c("lightslateblue","skyblue"))
myBlues <- b(3)

#party night
r <- colorRampPalette(c('firebrick1','salmon1'))
myReds <- r(4)
```


I also want to show  a link for the source data as an annotation
```{r}
a <- list(
  x = -0.05,
  y = -0.1,
  text = "Source: <a href='https://www.ubs.com/microsites/prices-earnings/en/stories/7-dinner-in-paris-party-in-miami-the-cheapest-and-priciest-cities-for-a-night-out/'>UBS</a>",
  xref = "paper",
  yref = "paper",
  showarrow = FALSE
)
```

Finally the call, adding title - and the plot! 
```{r}
p<- plot_ly() %>%
  add_bars(data= dataPlot,
          x = ~MyCost,
          y = ~City,
          group = ~Item,
          type = 'bar',
          showLegend = TRUE,
          text  = ~paste0(Item,": $",Cost),
          hoverinfo = 'text',
          orientation = 'h',
          marker = list(
            color = factor(dataPlot$Item,
                    labels=c(myBlues, myReds))))%>%
  layout(barmode = 'relative',
         xaxis = list(title= "Cost ($USD)",
                tickvals = tick_vals, ticktext= tick_labels),
         yaxis = list(title = ""),
         title = 'Makeover Monday 48: Cost of a night out around the world',
         margin = list(l = 100, r = 20, t = 70, b = 70),
         paper_bgcolor = 'rgb(248, 248, 255)',
         plot_bgcolor = 'rgb(248, 248, 255)',
         annotations = a,
         width = 650, 
         height = 500)

```

Voila !

```{r, echo=F}
p
#api_create(p, filename = "r-makeoverMonday-201848-costNightOut")
```

