---
title: 'Makeover Monday: Diversity in Puzzle constructors'
author: Felipe Marin
date: '2018-12-03'
slug: makeover-monday-diversity-in-puzzle-constructors
categories:
  - Visualisation
tags:
  [R, plotly, MakeoverMonday]
description: ''
featured_image: ''
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment=NA, fig.width=9, fig.height=6, warning = FALSE, message=FALSE)
```

This week the challenge was to explore a dataset of women who contributed 
to the New York Times crossword along the years, during the tenure of
Will Shortz. 

In the original [article](https://www.xwordinfo.com/Women) it was palpable
the under-representation by female contributors: From 922 puzzle constructors only 223
were women. 

I load the following packages:
```{r}
#load libraries
library(plotly)
library(readr)
library(dplyr)
library(magrittr)
library(data.world) #where data lives
```


```{r, echo=FALSE}

dw_womenCons <- "https://data.world/makeovermonday/2018w49/"

raw_data <- data.world::query(data.world::qry_sql("SELECT * 
                                            FROM women_constructors_in_the_shortz_era"),
                        dataset = dw_womenCons)

```


I create a custom function for the animation
```{r}
accumulate_by <- function(dat, var) {
  var <- lazyeval::f_eval(var, dat)
  lvls <- plotly:::getLevels(var)
  dats <- lapply(seq_along(lvls), function(x) {
  cbind(dat[var %in% lvls[seq(1, x)], ], frame = lvls[[x]])
  })
  dplyr::bind_rows(dats)
}
```

Then I add the percentage of women for each day per year
```{r}
#let's get percentage of women
dataPlot <- raw_data %>% 
  mutate(percWomen = round(women/(women+men)*100,2),
         percMen = round(men/(women+men)*100,2)) 

dataPlot$weekday <- factor(dataPlot$weekday,
              levels = c('Mon','Tue','Wed','Thu','Fri','Sat','Sun'))
dataPlot %<>% 
  mutate(nWeek = as.numeric(weekday))

#stats per year
dataYear <- dataPlot %>% 
  mutate(yeardate  = year) %>%
  group_by(yeardate) %>%
  summarise(percWomen = mean(percWomen, na.rm=T)) %>%
  accumulate_by(~yeardate) %>%
  mutate(year=frame)
  
```


The code below build the two side-to-side plots and the bar
```{r}
basePlot <- dataPlot %>%
  plot_ly(x = ~nWeek, y = ~percWomen,
          text = ~paste0("total female constructors ",year,": ",women), 
          hoverinfo = "text") 

tick_vals = c(1,2,3,4,5,6,7)
tick_labels = c('Mon','Tue','Wed','Thu','Fri','Sat','Sun')

#barplot 
p1 <- basePlot %>%
  add_bars( frame = ~year) %>%
  layout(xaxis = list(range=c(0.4,7.6),
                      tickvals = tick_vals, 
                      ticktext= tick_labels,
                      title = 'day of week (1-7: Mon-Sun)'),
      yaxis = list(title = "% of female constructors"),
         margin = list(t=70)) %>%
  add_annotations(
    text = "Women crosswords constructors in the NYT's Shortz Era - from <a href='https://www.xwordinfo.com/Women'>XWordInfo</a>",
    x = 0,
    y = 1.1,
    yref = "paper",
    xref = "paper",
    xanchor = "left",
    yanchor = "top",
    showarrow = FALSE,
    font = list(size = 18)
  )

#yearly plot
p2 <- plot_ly(dataYear) %>%
  add_lines(x=~yeardate, y=~percWomen, frame = ~year,color=I('orange'),
            text = ~paste0("weekly average: ",round(percWomen,1),"%"),
            hoverinfo='text') %>%
  layout(xaxis = list(range=c(1993,2017),title="year"),
         margin = list(t = 70),
         paper_bgcolor = 'rgb(248, 248, 255)',
         plot_bgcolor = 'rgb(248, 248, 255)') 
  
#animation slider
pp <- subplot(p1,p2,nrows=1,widths=c(0.7,0.3), titleX=TRUE,shareY = T,shareX=F) %>%
  hide_legend() %>%
 animation_opts(frame=1000,transition = 0) %>%
  animation_slider(
    currentvalue = list(prefix = "year ", font = list(color="gray"))
  )


pp
```


